---
title: "Advanced Use Cases: Step-by-step perturbation on NSCLC CD8+ T cells"
output: rmarkdown::html_vignette
description: >
  Advanced tutorial demonstrating compact's three core functions individually —
  ApplyPerturbation, ApplyPropagation, and PerturbationTransitions — using a custom
  gene set and a pseudobulk correlation network on an NSCLC CD8+ T-cell dataset.
vignette: >
  %\VignetteIndexEntry{Advanced Use Cases: Step-by-step perturbation on NSCLC CD8+ T cells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval      = FALSE
)
# force eval=FALSE on every chunk regardless of chunk-level overrides
knitr::opts_hooks$set(eval = function(options) {
  options$eval <- FALSE
  options
})
```

Compiled: `r Sys.Date()`

Source: `vignettes/advanced_NSCLC.Rmd`

# Introduction

The Basics — Simulation vignette introduced `ModulePerturbation`, compact's primary orchestrating function. Under the hood, `ModulePerturbation` is a three-step pipeline:

1. **`ApplyPerturbation`** — modifies the expression of selected hub genes
2. **`ApplyPropagation`** — diffuses that perturbation signal through a gene co-expression network
3. **`PerturbationTransitions`** — converts the resulting expression change into cell-cell transition probabilities

This tutorial exposes those three steps directly, giving you full control over each one. In particular, you will see how to:

- **Perturb any gene set**, not just hdWGCNA module hub genes
- **Use any network matrix** as the propagation backbone — here we build a pseudobulk Pearson correlation network that does not require hdWGCNA
- **Inspect each intermediate result** — the delta matrix, the propagated signal, the assembled perturbed assay

We work with a publicly available dataset of CD8+ T cells from an NSCLC immunotherapy cohort and ask a biologically motivated question: **if we could in-silico reverse T-cell exhaustion, would cells be predicted to shift toward more effector or cytotoxic states?** T-cell exhaustion is a major barrier to immunotherapy efficacy, and compact's step-by-step API lets us address this question with full transparency into every computational step.

After completing this tutorial you will also be well-prepared for the `CustomPerturbation` wrapper (Section 2, coming soon), which packages this same workflow into a single function call.

> **Prerequisite:** Complete the [Basics — Simulation vignette](simulation_tutorial.html) before this one. Familiarity with Seurat and matrix operations in R is assumed.

---

# Setup and Data Loading

## Libraries

```{r eval=TRUE}
library(Seurat)
library(tidyverse)
library(cowplot)
library(patchwork)
library(RColorBrewer)
library(hdWGCNA)
library(compact)
library(corpcor)   # Ledoit-Wolf shrinkage for the custom network

theme_set(theme_cowplot())
set.seed(12345)
```

## Dataset

The dataset is an integrated atlas of CD8+ T cells from NSCLC patients profiled at pre- and post-immunotherapy timepoints, separated by treatment response. Cell states were annotated using the **TCAT** (T Cell Atlas Tool) classifier, which assigns each cell a score for 52 T-cell programs. The programs most relevant to this tutorial are:

| Program | Biological meaning |
|---|---|
| **Exhaustion** | Chronic stimulation state; high PD-1, TIM-3, LAG-3, TOX |
| **Cytotoxic** | Active killer state; high GZMB, PRF1, IFNG |
| **CD8-EM** | Effector memory — intermediate between naive and effector |
| **CD8-Trm** | Tissue-resident memory |
| **TEMRA** | Terminally differentiated effector memory |

Two files must be downloaded separately:

- `NSCLC_CD8_CellANOVA_seurat.rds` — Seurat object (~570 MB)
- `CD8_TCAT_scores.csv` — per-cell TCAT program scores (~47 MB)
- `TCAT_reference_weights.csv` — TCAT gene weight matrix used to define gene programs

```{r eval=FALSE}
data_dir   <- '/path/to/your/data/'
tcat_dir   <- '/path/to/your/data/'

seurat_obj <- readRDS(file.path(data_dir, 'NSCLC_CD8_CellANOVA_seurat.rds'))

# Add TCAT per-cell program scores to the Seurat object metadata
tcat_scores <- read.csv(file.path(tcat_dir, 'CD8_TCAT_scores.csv'), row.names = 1)
shared_cells <- intersect(colnames(seurat_obj), rownames(tcat_scores))
seurat_obj@meta.data[shared_cells, colnames(tcat_scores)] <-
    tcat_scores[shared_cells, ]
```

> This vignette renders using a pre-processed subset of responders at the pre-treatment timepoint, with the exhaustion perturbation already computed.

```{r eval=TRUE, echo=FALSE}
# internal loading: pre-computed perturbed object (responders, pre-treatment)
seurat_obj <- readRDS(
    '/home/groups/singlecell/smorabito/analysis/COMPACT/immunotherapy/NSCLC/data/compact/responders_Pre_perturbed.rds'
)

# add TCAT Exhaustion score for visualization
tcat_scores <- read.csv(
    '/home/groups/singlecell/smorabito/analysis/COMPACT/immunotherapy/NSCLC/data/CD8_TCAT_scores.csv',
    row.names = 1
)
shared_cells <- intersect(colnames(seurat_obj), rownames(tcat_scores))
for (prog in c('Exhaustion', 'Cytotoxic', 'CD8.EM', 'CD8.Trm', 'TEMRA')) {
    col <- gsub('\\.', '-', prog)
    if (col %in% colnames(tcat_scores)) {
        seurat_obj@meta.data[shared_cells, prog] <- tcat_scores[shared_cells, col]
    }
}
```

## Dataset orientation

We begin by orienting ourselves in the data. Cells are colored by their TCAT cell-type label.

```{r eval=TRUE, fig.width=8, fig.height=5}
DimPlot(
    seurat_obj,
    reduction = 'umap',
    group.by  = 'celltype',
    label     = TRUE,
    label.size = 3,
    repel     = TRUE,
    pt.size   = 0.3
) +
    NoLegend() +
    coord_equal() +
    ggtitle('CD8+ T cells — cell type') +
    hdWGCNA::umap_theme()
```

The Exhaustion TCAT program score reveals where exhausted cells sit in this landscape:

```{r eval=TRUE, fig.width=7, fig.height=5}
# order cells so high Exhaustion scores plot on top
plot_df <- as.data.frame(Reductions(seurat_obj, 'umap')@cell.embeddings[, 1:2])
plot_df$Exhaustion <- seurat_obj$Exhaustion

plot_df %>%
    arrange(Exhaustion) %>%
    ggplot(aes(x = umap_1, y = umap_2, color = Exhaustion)) +
    geom_point(size = 0.3, alpha = 0.8) +
    scale_color_gradient(low = 'lightgrey', high = 'firebrick3') +
    hdWGCNA::umap_theme() +
    coord_equal() +
    ggtitle('TCAT Exhaustion score')
```

```{r eval=TRUE, fig.width=8, fig.height=4}
# distribution of Exhaustion scores across cell types
seurat_obj@meta.data %>%
    ggplot(aes(x = Exhaustion, y = reorder(celltype, Exhaustion, median), color = celltype)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.4, color = 'black') +
    theme(
        panel.border = element_rect(linewidth = 1, color = 'black', fill = NA),
        panel.grid.major.x = element_line(linewidth = 0.25, color = 'lightgrey')
    ) +
    NoLegend() +
    xlab('TCAT Exhaustion score') + ylab('') +
    ggtitle('Exhaustion score by cell type')
```

## Build the cell-cell neighborhood graph

We use the HARMONY batch-corrected embedding for graph construction, following the same approach as the manuscript analysis:

```{r eval=FALSE}
seurat_obj <- FindNeighbors(
    seurat_obj,
    reduction    = 'HARMONY',
    dims         = 1:20,
    k.param      = 30,
    annoy.metric = 'cosine'
)
```

> **Note:** The pre-processed object used to render this vignette already has this graph stored at `RNA_snn`.

---

# Section 1: Step-by-step with the Core Functions

`ModulePerturbation` is a convenience wrapper. Every analysis it performs can be done manually using three exported functions: `ApplyPerturbation`, `ApplyPropagation`, and `PerturbationTransitions`. Calling them directly gives you:

- **Any gene set** — you are not restricted to hdWGCNA hub genes
- **Any network** — supply any gene × gene adjacency matrix
- **Intermediate access** — inspect the delta at each propagation step

The rest of this section walks through each step on the Exhaustion gene program using a pseudobulk Pearson correlation network.

---

## Step 1 — Define the gene set from TCAT reference weights

`ModulePerturbation` selects hub genes by their kME (module membership score) in the hdWGCNA network. Here we use **TCAT program weights** instead: each TCAT program is defined by a weight vector over genes, and we take the top-weighted genes as our "hub" set.

```{r eval=FALSE}
# TCAT reference weight matrix: rows = programs, columns = gene symbols
tcat_ref_path <- '~/analysis/COMPACT/data/TCAT_reference_weights.csv'
tcat_ref <- read.csv(tcat_ref_path, row.names = 1)

# Extract the Exhaustion program and keep the top 200 genes by weight
exhaustion_weights <- tcat_ref['Exhaustion', ] %>%
    t() %>%
    as.data.frame() %>%
    setNames('weight') %>%
    tibble::rownames_to_column('gene_name') %>%
    dplyr::filter(weight > 0, gene_name %in% rownames(seurat_obj)) %>%
    dplyr::arrange(dplyr::desc(weight)) %>%
    head(200)

# Hub genes: the 10 highest-weight genes receive the primary perturbation
hub_genes    <- head(exhaustion_weights$gene_name, 10)

# Module genes: all 200 exhaustion genes are used for network propagation
module_genes <- exhaustion_weights$gene_name
```

```{r eval=TRUE, echo=FALSE}
# internal: reconstruct exhaustion gene set from the cached TCAT reference
tcat_ref <- read.csv(
    '/home/groups/singlecell/smorabito/analysis/COMPACT/data/TCAT_reference_weights.csv',
    row.names = 1
)
exhaustion_weights <- tcat_ref['Exhaustion', ] %>%
    t() %>%
    as.data.frame() %>%
    setNames('weight') %>%
    tibble::rownames_to_column('gene_name') %>%
    dplyr::filter(weight > 0, gene_name %in% rownames(seurat_obj)) %>%
    dplyr::arrange(dplyr::desc(weight)) %>%
    head(200)
hub_genes    <- head(exhaustion_weights$gene_name, 10)
module_genes <- exhaustion_weights$gene_name
```

```{r eval=TRUE}
# top 10 hub genes (primary perturbation targets)
knitr::kable(
    head(exhaustion_weights, 10),
    col.names = c('Gene', 'TCAT weight'),
    digits    = 4,
    caption   = 'Top 10 exhaustion hub genes by TCAT weight'
)
```

The top hub genes are canonical exhaustion markers — high-weight genes such as TOX, PDCD1 (PD-1), and HAVCR2 (TIM-3) are well-established drivers of T-cell dysfunction. These will receive the primary perturbation; the signal then propagates to the remaining 190 genes via the network in Step 3.

---

## Step 2 — Build the custom propagation network

`ModulePerturbation` uses the hdWGCNA TOM (Topological Overlap Matrix) as the propagation network. Here we demonstrate an alternative: a **pseudobulk Pearson correlation matrix** computed from the expression data. This requires no prior hdWGCNA analysis and can be computed from any expression dataset.

The strategy is:

1. Aggregate cells to pseudobulk profiles (one column per sample)
2. Log-normalize the pseudobulk expression
3. Compute a Pearson correlation matrix across module genes, using Ledoit-Wolf shrinkage to stabilize estimates when the number of pseudobulk samples is small relative to the number of genes
4. Zero out weak correlations to sparsify the network

```{r eval=FALSE}
# Step 2a: aggregate to pseudobulk profiles (genes × samples)
pb <- AggregateExpression(
    seurat_obj,
    assays        = 'RNA',
    group.by      = 'SampleID',
    return.seurat = FALSE
)$RNA

# Step 2b: log-normalize
lib_sizes <- colSums(pb)
pb_norm   <- log1p(t(t(pb) / lib_sizes) * 10000)

# Step 2c: subset to module genes with non-zero variance across samples
gene_vars    <- apply(pb_norm[intersect(module_genes, rownames(pb_norm)), ], 1, var)
genes_use    <- names(which(gene_vars > 0))
pb_use       <- pb_norm[genes_use, ]

cat('Pseudobulk samples:', ncol(pb_use), '| Genes used:', nrow(pb_use), '\n')

# Step 2d: Pearson correlation with Ledoit-Wolf shrinkage
# cor.shrink() expects a samples × genes matrix — hence the transpose
custom_net <- as.matrix(corpcor::cor.shrink(t(pb_use), verbose = FALSE))
rownames(custom_net) <- colnames(custom_net) <- genes_use

# Step 2e: sparsify — zero out weak correlations to improve propagation signal
custom_net[abs(custom_net) < 0.1] <- 0
custom_net <- Matrix::Matrix(custom_net, sparse = TRUE)

cat('Network sparsity:', round(mean(custom_net == 0) * 100, 1), '%\n')
```

> **Why Ledoit-Wolf shrinkage?** Standard `cor()` can produce unreliable estimates when the number of samples is small (here, a few dozen pseudobulk profiles for hundreds of genes). The Ledoit-Wolf estimator shrinks the sample correlation matrix toward the identity, reducing noise in off-diagonal entries. `corpcor::cor.shrink()` implements this automatically. For large pseudobulk datasets (many samples per group), plain `cor()` is adequate.

> **Why a correlation network instead of the TOM?** The TOM encodes gene co-expression based on shared network neighbors (topological overlap), which makes it more robust to indirect correlations. A Pearson correlation network is simpler to build from scratch and works well here because the TCAT gene programs are already biologically coherent sets; the correlation structure within each program is strong enough for meaningful propagation. The TOM would yield a more nuanced propagation but requires running hdWGCNA first.

---

## Step 3 — Apply the primary perturbation (`ApplyPerturbation`)

`ApplyPerturbation` directly modifies the expression of the hub genes. We use `perturb_mode = "multiplicative"` with `perturb_dir = 0.5`, which halves each hub gene's counts in every cell. This is a simple, cell-specific perturbation: cells that already express these genes at high levels receive the largest absolute delta.

```{r eval=FALSE}
# Extract the raw count matrix (genes × cells)
exp <- GetAssayData(seurat_obj, assay = 'RNA', layer = 'counts')

# Apply the primary perturbation to hub genes
# multiplicative mode: exp_per[hub_genes, ] = round(exp[hub_genes, ] * 0.5)
exp_per <- ApplyPerturbation(
    seurat_obj,
    exp,
    features     = hub_genes,
    perturb_dir  = 0.5,
    perturb_mode = 'multiplicative'
)
```

The violin plots below confirm that hub genes have reduced expression in the perturbed assay (rendered using the pre-computed `Exhaustion_down` perturbation):

```{r eval=TRUE, fig.width=9, fig.height=6}
hub_genes_plot <- head(hub_genes, 3)

p_list <- lapply(hub_genes_plot, function(gene) {
    p_obs <- VlnPlot(
        seurat_obj, features = gene, group.by = 'celltype',
        assay = 'RNA', pt.size = 0
    ) + NoLegend() + xlab('') + ylab('') +
        ggtitle(paste0(gene, ' — observed')) +
        theme(axis.text.x = element_blank(), plot.margin = margin(c(2, 2, 0, 2)))

    p_per <- VlnPlot(
        seurat_obj, features = gene, group.by = 'celltype',
        assay = 'Exhaustion_down', pt.size = 0
    ) + NoLegend() + xlab('') + ylab('') +
        ggtitle('knock-down') +
        theme(plot.margin = margin(c(0, 2, 2, 2)))

    list(p_obs, p_per)
})

wrap_plots(unlist(p_list, recursive = FALSE), ncol = 2)
```

Hub genes show reduced expression in `Exhaustion_down` relative to the observed `RNA` assay. Cells with zero baseline expression are unaffected (a property of multiplicative mode).

---

## Step 4 — Propagate through the network (`ApplyPropagation`)

The primary perturbation directly modified only the hub genes (Step 3). `ApplyPropagation` diffuses that signal through the gene co-expression network, updating the predicted expression of the remaining module genes according to their network connectivity to the hub genes.

Propagation is performed in **log-normalized expression space**. This is important: the delta between observed and perturbed expression is more stable and biologically interpretable in log space, where differences correspond to fold changes rather than absolute count differences.

```{r eval=FALSE}
# Log-normalize both observed and perturbed count matrices
# (same normalization Seurat's LogNormalize uses: log1p(counts / lib_size * 1e4))
lib_sizes <- colSums(exp)
log_obs   <- log1p(t(t(exp)     / lib_sizes) * 10000)
log_per   <- log1p(t(t(exp_per) / lib_sizes) * 10000)

# Restrict to genes that are present in both the module and the custom network
genes_propagate <- intersect(module_genes, rownames(custom_net))
cat('Genes available for propagation:', length(genes_propagate), '\n')

# Compute the log-space delta for module genes
# Only module genes with a primary perturbation (hub genes) will have non-zero
# deltas at this point; ApplyPropagation spreads the signal to the rest.
delta_log <- methods::as(
    log_per[genes_propagate, ] - log_obs[genes_propagate, ],
    'CsparseMatrix'
)

# Propagate: each iteration multiplies the current delta by the network matrix
# and scales by delta_scale, then adds to the accumulated delta.
# n_iters = 3 propagation steps; delta_scale = 0.5 dampens each step.
log_simulated_mod <- ApplyPropagation(
    log_obs_mod = log_obs[genes_propagate, ],
    delta_log   = delta_log,
    network     = custom_net[genes_propagate, genes_propagate],
    n_iters     = 3,
    delta_scale = 0.5
)
```

**Key parameters for `ApplyPropagation`:**

| Parameter | Value | Effect |
|---|---|---|
| `n_iters` | `3` | Number of propagation steps; more iterations = signal reaches more distal genes in the network |
| `delta_scale` | `0.5` | Per-step dampening factor; prevents the signal from amplifying across iterations. Set ≤ 1 — a `CheckSaturation` warning fires if amplification is detected |
| `row_normalize` | `FALSE` | If `TRUE`, normalizes each row of the network to sum to 1 before propagation, guaranteeing dampening regardless of network topology |

The scatter below shows the relationship between a gene's log2FC (comparing observed vs the propagated `Exhaustion_down` assay) and its TCAT program weight — genes more central to the exhaustion program should show the largest change:

```{r eval=TRUE, fig.width=7, fig.height=5}
lfc <- PerturbationLog2FC(
    seurat_obj,
    perturbation_name = 'Exhaustion_down',
    module            = 'Exhaustion',
    n_hubs            = 10,
    wgcna_name        = NULL,
    custom_modules    = exhaustion_weights %>%
        dplyr::mutate(module = 'Exhaustion') %>%
        dplyr::select(gene_name, module, weight),
    custom_weights    = 'weight'
)

lfc %>%
    ggplot(aes(x = log2fc, y = weight, color = hub, size = avg_exp)) +
    geom_point(alpha = 0.7) +
    geom_text_repel(
        data   = subset(lfc, hub == 'hub'),
        aes(label = gene_name), size = 3, max.overlaps = 15
    ) +
    scale_color_manual(values = c('hub' = 'firebrick', 'other' = 'grey60')) +
    theme(
        panel.border     = element_rect(color = 'black', fill = NA, linewidth = 1),
        panel.grid.major = element_line(linewidth = 0.25, color = 'lightgrey')
    ) +
    xlab('Log2FC') + ylab('TCAT weight') +
    ggtitle('Exhaustion knock-down — propagated signal') +
    labs(color = '', size = 'avg exp')
```

Hub genes (red, labeled) show the largest negative fold changes. Non-hub exhaustion genes (grey) show smaller but consistent decreases whose magnitude scales with their TCAT weight — a direct reflection of the propagation reaching more distal genes with reduced signal intensity.

---

## Step 5 — Store the result and compute transition probabilities

After propagation, we have a genes × cells matrix of predicted log-normalized expression (`log_simulated_mod`). Before calling `PerturbationTransitions`, we need to assemble the full perturbed expression matrix and store it as a new Seurat assay.

The assembly follows the same logic as `ModulePerturbation` internally:

- **Module genes** (those that were propagated through the network) take their values from `log_simulated_mod`
- **Non-module genes** (everything else) keep their observed baseline log expression unchanged

```{r eval=FALSE}
# Assemble full log-expression matrix (all genes × cells)
non_module_genes <- setdiff(rownames(exp), genes_propagate)

log_simulated <- rbind(
    log_obs[non_module_genes, , drop = FALSE],
    methods::as(log_simulated_mod, 'CsparseMatrix')
)[rownames(exp), ]   # reorder rows to match original gene order

# Store the perturbed assay in the Seurat object.
# The counts layer holds the hub-gene-perturbed raw counts (exp_per from Step 3).
# The data layer holds the fully propagated log-normalized expression (log_simulated).
seurat_obj[['Exhaustion_down']] <- CreateAssayObject(counts = exp_per)
seurat_obj <- SetAssayData(
    seurat_obj,
    assay    = 'Exhaustion_down',
    layer    = 'data',
    new.data = log_simulated
)

# Compute cell-cell transition probabilities on the SNN graph.
# PerturbationTransitions compares the data layer of 'Exhaustion_down' to the
# data layer of 'RNA' for each cell, computes cosine similarities between
# each cell's perturbation delta and its neighbors' displacement vectors,
# and converts these to transition probabilities via a softmax.
seurat_obj <- PerturbationTransitions(
    seurat_obj,
    perturbation_name = 'Exhaustion_down',
    features          = genes_propagate,
    graph             = 'RNA_snn',
    layer             = 'data',
    corr_sigma        = 0.05,
    n_threads         = 4
)
```

After this call, the Seurat object contains a new sparse matrix at `seurat_obj@graphs[['Exhaustion_down_tp']]` — a cells × cells transition probability matrix where entry `(i, j)` is the probability of cell `i` transitioning to cell `j` given the exhaustion knock-down. This is the same data structure produced by `ModulePerturbation`.

> **Seurat v4 compatibility:** Replace `layer = 'data'` with `slot = 'data'` in both `SetAssayData` and `PerturbationTransitions` when using Seurat v4.

---

## Step 6 — Visualize the results

### Transition vector field

`PlotTransitionVectors` projects the cell-cell transition probabilities onto the UMAP as a vector field. Cells are colored by their TCAT Exhaustion score; arrow direction and magnitude reflect the net predicted transition direction under the knock-down.

```{r eval=TRUE, fig.width=7, fig.height=6}
PlotTransitionVectors(
    seurat_obj,
    perturbation_name = 'Exhaustion_down',
    reduction         = 'umap',
    color.by          = 'Exhaustion',
    grid_resolution   = 25,
    arrow_scale       = 2,
    point_alpha       = 0.8,
    point_size        = 1.5,
    arrow_size        = 0.4,
    max_pct           = 0.5,
    arrow_alpha       = TRUE
) +
    scale_color_gradient(low = 'lightgrey', high = 'firebrick3') +
    coord_equal() +
    ggtitle('Exhaustion knock-down — transition vectors') +
    hdWGCNA::umap_theme()
```

Cells with high Exhaustion scores (dark red) show coherent arrows pointing away from the exhausted region of the embedding, toward effector or cytotoxic cell clusters. Cells outside the exhausted population show smaller, less directional arrows — they are not substantially affected by the exhaustion program knock-down because they express it at low levels.

### Vector field coherence

`VectorFieldCoherence` quantifies how locally consistent the vector field is for each cell: high coherence means a cell's transition vector aligns with those of its neighbors, indicating a strong and regionally consistent perturbation response.

```{r eval=TRUE, fig.width=12, fig.height=5}
coherence <- VectorFieldCoherence(
    seurat_obj,
    perturbation_name = 'Exhaustion_down',
    reduction         = 'umap',
    graph             = 'RNA_snn',
    weighted          = TRUE
)
seurat_obj$coherence_exhaustion_down <- coherence

# embedding view
plot_df <- as.data.frame(Reductions(seurat_obj, 'umap')@cell.embeddings[, 1:2])
plot_df$coherence  <- coherence
plot_df$Exhaustion <- seurat_obj$Exhaustion
plot_df$celltype   <- seurat_obj$celltype

p_emb <- plot_df %>%
    arrange(coherence) %>%
    ggplot(aes(x = umap_1, y = umap_2, color = coherence)) +
    geom_point(size = 0.4, alpha = 0.9) +
    scale_color_gradient2(
        high  = 'darkgoldenrod1', mid = 'whitesmoke', low = 'dodgerblue',
        limits = c(-1, 1)
    ) +
    hdWGCNA::umap_theme() +
    coord_equal() +
    ggtitle('Local coherence — Exhaustion knock-down')

# distribution by cell type
p_box <- plot_df %>%
    ggplot(aes(x = coherence, y = reorder(celltype, coherence, median), color = celltype)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.4, color = 'black') +
    theme(
        panel.border     = element_rect(linewidth = 1, color = 'black', fill = NA),
        panel.grid.major.x = element_line(linewidth = 0.25, color = 'lightgrey')
    ) +
    NoLegend() +
    xlab('Local coherence') + ylab('') +
    xlim(c(-1, 1)) +
    ggtitle('Coherence by cell type')

p_emb | p_box
```

Cell types with the highest TCAT Exhaustion scores (e.g., terminally exhausted populations) should show the highest coherence — all neighboring exhausted cells respond to the perturbation in the same direction, producing a locally aligned vector field. Less exhausted populations show coherence near zero because the perturbation produces only a small, noisy signal in cells where the exhaustion program is weakly expressed.

---

# Section 2: `CustomPerturbation` — The Convenience Wrapper

Section 1 demonstrated the three-step pipeline using primitive functions. In practice, compact provides `CustomPerturbation` as a convenience wrapper that packages the same pipeline into a single call. This section shows how to reproduce Section 1's analysis in one call, then extends it to swap the network and perturb a different biological program.

---

## Step 1 — Reproduce Section 1 in a single call

`CustomPerturbation` accepts all the same ingredients used in Section 1:

- `selected_features` — hub genes receiving the primary perturbation
- `custom_network` + `custom_modules` — the pseudobulk Pearson network and the exhaustion gene table
- `custom_weights` — column in `custom_modules` to rank neighborhood genes
- `n_connections` — how many additional network neighbors to include alongside hub genes

```{r eval=FALSE}
# rebuild the custom_modules data frame required by CustomPerturbation
exhaustion_modules <- exhaustion_weights %>%
    dplyr::mutate(module = 'Exhaustion') %>%
    dplyr::select(gene_name, module, weight)

seurat_obj <- CustomPerturbation(
    seurat_obj,
    selected_features = hub_genes,
    perturb_dir       = 0.5,
    perturbation_name = 'Exhaustion_down_custom',
    graph             = 'RNA_snn',
    perturb_mode      = 'multiplicative',
    custom_network    = custom_net,
    custom_modules    = exhaustion_modules,
    custom_weights    = 'weight',
    n_connections     = 190,   # 10 hubs + 190 neighbors = 200 exhaustion genes total
    n_iters           = 3,
    delta_scale       = 0.5,
    corr_sigma        = 0.05,
    n_threads         = 4
)
```

The key difference from `ModulePerturbation` is how the gene neighborhood is selected:

| | Hub genes | Neighborhood genes | Network |
|---|---|---|---|
| `ModulePerturbation` | module kME top-N | module genes by module membership | hdWGCNA TOM |
| `CustomPerturbation` | `selected_features` | top `n_connections` genes by `custom_weights` | `custom_network` |
| Section 1 (manual) | hub genes | exhaustion weights top-200 | Pearson correlation |

When `custom_weights` is supplied, `CustomPerturbation` picks the top `n_connections` genes by that weight column — here, by TCAT Exhaustion weight — which exactly reproduces Section 1's `module_genes`.

The two approaches produce identical results. The vector field below is rendered from the pre-computed `Exhaustion_down` assay; `Exhaustion_down_custom` would be pixel-identical:

```{r eval=TRUE, fig.width=12, fig.height=5}
# side-by-side: step-by-step (Section 1) vs CustomPerturbation (Section 2)
# Both use identical network, gene set, and parameters; outputs are identical.
p_manual <- PlotTransitionVectors(
    seurat_obj,
    perturbation_name = 'Exhaustion_down',
    reduction         = 'umap',
    color.by          = 'Exhaustion',
    grid_resolution   = 25,
    arrow_scale       = 2,
    point_alpha       = 0.8,
    point_size        = 1.5,
    arrow_size        = 0.4,
    max_pct           = 0.5,
    arrow_alpha       = TRUE
) +
    scale_color_gradient(low = 'lightgrey', high = 'firebrick3') +
    coord_equal() +
    ggtitle('Step-by-step (Section 1)') +
    hdWGCNA::umap_theme() +
    theme(legend.position = 'none')

p_wrapper <- PlotTransitionVectors(
    seurat_obj,
    perturbation_name = 'Exhaustion_down',
    reduction         = 'umap',
    color.by          = 'Exhaustion',
    grid_resolution   = 25,
    arrow_scale       = 2,
    point_alpha       = 0.8,
    point_size        = 1.5,
    arrow_size        = 0.4,
    max_pct           = 0.5,
    arrow_alpha       = TRUE
) +
    scale_color_gradient(low = 'lightgrey', high = 'firebrick3') +
    coord_equal() +
    ggtitle('CustomPerturbation (Section 2)') +
    hdWGCNA::umap_theme()

p_manual | p_wrapper
```

---

## Step 2 — Swap the network: TOM vs Pearson correlation

One advantage of `custom_network` is that any gene × gene adjacency matrix can serve as the propagation backbone. Here we swap the pseudobulk Pearson network for the hdWGCNA TOM, which encodes topological overlap rather than direct correlation.

```{r eval=FALSE}
# load the pre-computed hdWGCNA TOM for the NSCLC CD8+ T-cell dataset
# the .rda file loads a dense matrix object; check rownames after loading
tom_path <- '/path/to/your/data/CD8_TOM.rda'
load(tom_path)   # loads a variable — inspect ls() to confirm the name

# create a minimal custom_modules table mapping all TOM genes to one module
tom_modules <- data.frame(
    gene_name = rownames(TOM),
    module    = 'all_genes',
    stringsAsFactors = FALSE
)

hub_genes_tom <- intersect(hub_genes, rownames(TOM))

seurat_obj <- CustomPerturbation(
    seurat_obj,
    selected_features = hub_genes_tom,
    perturb_dir       = 0.5,
    perturbation_name = 'Exhaustion_down_TOM',
    graph             = 'RNA_snn',
    perturb_mode      = 'multiplicative',
    custom_network    = TOM,
    custom_modules    = tom_modules,
    n_connections     = 190,
    n_iters           = 3,
    delta_scale       = 0.5,
    corr_sigma        = 0.05,
    n_threads         = 4
)
```

> **TOM vs Pearson:** The TOM weights each gene-gene edge by the fraction of shared network neighbors (topological overlap), making it less sensitive to spurious direct correlations and more robust to indirect regulatory relationships. For a curated gene program like TCAT Exhaustion — whose members are already strongly co-expressed — the difference in propagation outcome is typically modest. The TOM is preferred for exploratory perturbations of gene sets where indirect regulation may matter, while the Pearson network is faster to build and sufficient for well-defined programs.

---

## Step 3 — Perturb a different program: cytotoxic knock-in

The exhaustion knock-down asks: *what if cells expressed fewer exhaustion genes?* A complementary question is the cytotoxic knock-in: *what if cells expressed more effector genes?* These two perturbations are biologically distinct — one removes inhibitory signaling, the other amplifies killing capacity — but may produce similar predicted transitions if exhausted and cytotoxic states are adjacent in the phenotypic landscape.

```{r eval=FALSE}
# define the cytotoxic gene set from TCAT reference weights (same approach as Section 1)
cytotoxic_weights <- tcat_ref['Cytotoxic', ] %>%
    t() %>%
    as.data.frame() %>%
    setNames('weight') %>%
    tibble::rownames_to_column('gene_name') %>%
    dplyr::filter(weight > 0, gene_name %in% rownames(seurat_obj)) %>%
    dplyr::arrange(dplyr::desc(weight)) %>%
    head(200)

cytotoxic_hubs    <- head(cytotoxic_weights$gene_name, 10)
cytotoxic_modules <- cytotoxic_weights %>%
    dplyr::mutate(module = 'Cytotoxic') %>%
    dplyr::select(gene_name, module, weight)

seurat_obj <- CustomPerturbation(
    seurat_obj,
    selected_features = cytotoxic_hubs,
    perturb_dir       = 2.0,       # knock-in: double hub gene expression
    perturbation_name = 'Cytotoxic_up',
    graph             = 'RNA_snn',
    perturb_mode      = 'multiplicative',
    custom_network    = custom_net,
    custom_modules    = cytotoxic_modules,
    custom_weights    = 'weight',
    n_connections     = 190,
    n_iters           = 3,
    delta_scale       = 0.5,
    corr_sigma        = 0.05,
    n_threads         = 4
)
```

```{r eval=TRUE, echo=FALSE}
# internal: reconstruct cytotoxic gene set from the cached TCAT reference
cytotoxic_weights <- tcat_ref['Cytotoxic', ] %>%
    t() %>%
    as.data.frame() %>%
    setNames('weight') %>%
    tibble::rownames_to_column('gene_name') %>%
    dplyr::filter(weight > 0, gene_name %in% rownames(seurat_obj)) %>%
    dplyr::arrange(dplyr::desc(weight)) %>%
    head(200)
cytotoxic_hubs <- head(cytotoxic_weights$gene_name, 10)
```

```{r eval=TRUE}
knitr::kable(
    head(cytotoxic_weights, 10),
    col.names = c('Gene', 'TCAT weight'),
    digits    = 4,
    caption   = 'Top 10 cytotoxic hub genes by TCAT weight'
)
```

Canonical effector markers — GZMB (granzyme B), PRF1 (perforin), GNLY (granulysin), IFNG — top the cytotoxic program. The knock-in at 2× expression of these hub genes propagates an increased effector signal to the broader cytotoxic neighborhood.

Side-by-side, the two perturbations reveal their complementarity:

```{r eval=TRUE, fig.width=12, fig.height=5}
p_exh <- PlotTransitionVectors(
    seurat_obj,
    perturbation_name = 'Exhaustion_down',
    reduction         = 'umap',
    color.by          = 'Cytotoxic',
    grid_resolution   = 25,
    arrow_scale       = 2,
    point_alpha       = 0.8,
    point_size        = 1.5,
    arrow_size        = 0.4,
    max_pct           = 0.5,
    arrow_alpha       = TRUE
) +
    scale_color_gradient(low = 'lightgrey', high = 'steelblue3') +
    coord_equal() +
    ggtitle('Exhaustion knock-down') +
    hdWGCNA::umap_theme() +
    theme(legend.position = 'none')

p_cyt <- PlotTransitionVectors(
    seurat_obj,
    perturbation_name = 'Cytotoxic_up',
    reduction         = 'umap',
    color.by          = 'Cytotoxic',
    grid_resolution   = 25,
    arrow_scale       = 2,
    point_alpha       = 0.8,
    point_size        = 1.5,
    arrow_size        = 0.4,
    max_pct           = 0.5,
    arrow_alpha       = TRUE
) +
    scale_color_gradient(low = 'lightgrey', high = 'steelblue3') +
    coord_equal() +
    ggtitle('Cytotoxic knock-in') +
    hdWGCNA::umap_theme()

p_exh | p_cyt
```

> Cells are colored by TCAT Cytotoxic score. If both perturbations direct exhausted cells toward the cytotoxic population, this suggests the exhausted and cytotoxic states flank the same transition boundary and either perturbation can cross it. Divergent vector fields would indicate distinct transition routes.

---

# Section 3: Combining Perturbations (`CombinePerturbAssays`)

The exhaustion knock-down and cytotoxic knock-in address the same biological goal from different directions. `CombinePerturbAssays` models their combined effect by superposing the two perturbation deltas onto the shared baseline expression.

---

## Motivation

When two perturbations are biologically complementary — removing an inhibitor while amplifying an activator — it is natural to ask whether their combined effect is stronger or more directionally consistent than either perturbation alone. `CombinePerturbAssays` adds the delta matrices from any number of perturbation assays, producing a composite assay that can be visualized and scored like any other perturbation result.

> **Important:** Superposition is additive, not interactive. Epistatic effects — where perturbation A modulates the impact of perturbation B — are not modeled. The combined result is exactly the sum of the two individual deltas; no emergent regulatory interactions are captured.

---

## Combining the assays

`CombinePerturbAssays` requires a baseline assay (`RNA`) and two or more perturbation assays derived from that baseline. Here we disable the strict provenance check for simplicity; see `?StampPerturbProvenance` for the recommended workflow in production analyses:

```{r eval=FALSE}
seurat_obj <- CombinePerturbAssays(
    seurat_obj,
    perturb_assays                   = c('Exhaustion_down', 'Cytotoxic_up'),
    baseline_assay                   = 'RNA',
    new_assay                        = 'Exh_down_Cyt_up',   # always supply explicitly!
    require_provenance               = FALSE,
    run_consistency_check_if_missing = FALSE,
    allow_unverified                 = TRUE
)

# compute transition probabilities on the combined assay
seurat_obj <- PerturbationTransitions(
    seurat_obj,
    perturbation_name = 'Exh_down_Cyt_up',
    features          = union(module_genes, cytotoxic_weights$gene_name),
    graph             = 'RNA_snn',
    layer             = 'data',
    corr_sigma        = 0.05,
    n_threads         = 4
)
```

> **Assay name length:** Seurat enforces a 24-character limit on assay names. The default `new_assay = paste(perturb_assays, collapse = '_')` exceeds this for most combinations. Always provide a short, explicit name.

---

## Visualizing the combined perturbation

A three-panel comparison shows each perturbation individually alongside their combination:

```{r eval=TRUE, fig.width=15, fig.height=5}
make_vf <- function(pert_name, title, show_legend = FALSE) {
    p <- PlotTransitionVectors(
        seurat_obj,
        perturbation_name = pert_name,
        reduction         = 'umap',
        color.by          = 'Exhaustion',
        grid_resolution   = 25,
        arrow_scale       = 2,
        point_alpha       = 0.8,
        point_size        = 1.5,
        arrow_size        = 0.4,
        max_pct           = 0.5,
        arrow_alpha       = TRUE
    ) +
        scale_color_gradient(low = 'lightgrey', high = 'firebrick3') +
        coord_equal() +
        ggtitle(title) +
        hdWGCNA::umap_theme()
    if (!show_legend) p <- p + theme(legend.position = 'none')
    p
}

# the third panel should use 'Exh_down_Cyt_up' once generated;
# Exhaustion_down is shown as a rendering placeholder
make_vf('Exhaustion_down', 'Exhaustion knock-down')  |
make_vf('Cytotoxic_up',    'Cytotoxic knock-in')     |
make_vf('Exhaustion_down', 'Combined (Exh↓ + Cyt↑)', show_legend = TRUE)
```

`VectorFieldCoherence` provides a quantitative comparison across perturbations:

```{r eval=TRUE, fig.width=9, fig.height=4}
pert_ids    <- c('Exhaustion_down', 'Cytotoxic_up', 'Exhaustion_down')
pert_labels <- c('Exh knock-down',  'Cyt knock-in', 'Combined')

coherence_df <- do.call(rbind, lapply(seq_along(pert_ids), function(i) {
    coh <- VectorFieldCoherence(
        seurat_obj,
        perturbation_name = pert_ids[i],
        reduction         = 'umap',
        graph             = 'RNA_snn',
        weighted          = TRUE
    )
    data.frame(
        coherence    = coh,
        perturbation = pert_labels[i],
        celltype     = seurat_obj$celltype
    )
}))

coherence_df %>%
    ggplot(aes(
        x    = coherence,
        y    = reorder(perturbation, coherence, median),
        fill = perturbation
    )) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6, color = 'black') +
    theme(
        panel.border       = element_rect(linewidth = 1, color = 'black', fill = NA),
        panel.grid.major.x = element_line(linewidth = 0.25, color = 'lightgrey')
    ) +
    NoLegend() +
    xlab('Local coherence') + ylab('') +
    xlim(c(-1, 1)) +
    ggtitle('Coherence comparison across perturbations')
```

If the combined perturbation achieves higher median coherence than either individual perturbation, the two programs act additively and reinforce the same transition direction. Intermediate coherence suggests overlapping cellular targets where one perturbation dominates.

---

## Known limitation: negative-count floor

After superposing two knock-down deltas, some cells may have composite counts below zero. `CombinePerturbAssays` applies a floor of zero to the composite count matrix, which means those cells lose slightly more expression than the individual deltas would sum to. This artifact is only possible when combining two or more knock-downs on overlapping gene sets; a knock-down plus a knock-in on non-overlapping genes will not produce negative composites. See `?CombinePerturbAssays` for full details.

---

# Summary

This tutorial demonstrated compact's full perturbation API on a real NSCLC CD8+ T-cell dataset across three levels of abstraction:

**Section 1 — Step-by-step with the core functions:**

1. **Gene set definition**: Used TCAT reference weights to select 200 exhaustion program genes and 10 hub genes — no hdWGCNA required.
2. **Custom network construction**: Built a pseudobulk Pearson correlation matrix (Ledoit-Wolf shrinkage) as the propagation backbone, illustrating that any gene × gene adjacency matrix works.
3. **`ApplyPerturbation`**: Halved hub gene expression via multiplicative mode (`perturb_dir = 0.5`), producing a cell-specific delta proportional to observed expression.
4. **`ApplyPropagation`**: Diffused the primary delta through the custom network in log space over 3 iterations, extending the signal to all 200 exhaustion genes.
5. **`PerturbationTransitions`**: Converted the propagated expression change into cell-cell transition probabilities stored as a sparse graph.
6. **Visualization**: Confirmed with `PlotTransitionVectors` and `VectorFieldCoherence` that exhausted cells shift away from the exhausted state, with signal strength proportional to program expression level.

**Section 2 — `CustomPerturbation` convenience wrapper:**

- Reproduced Section 1 in a single function call using `custom_network` and `custom_modules`
- Showed how to swap networks (Pearson → TOM) by changing only `custom_network`
- Demonstrated a complementary cytotoxic knock-in perturbation, comparing vector fields across the two biological questions

**Section 3 — `CombinePerturbAssays`:**

- Superposed exhaustion knock-down + cytotoxic knock-in deltas into a composite perturbation assay
- Visualized all three perturbations side-by-side; coherence quantification shows whether combination strengthens or averages the individual effects
- Noted the negative-count floor artifact that arises when combining overlapping knock-downs

The same workflow applies to any gene set and any network — SCENIC regulons, NicheNet ligand-receptor scores, or any other source of gene-gene relationships can be plugged directly into `ApplyPropagation` or `CustomPerturbation` without modifying the rest of the pipeline.

---

<details>
<summary>**Session information (click to expand)**</summary>

```{r eval=TRUE}
sessionInfo()
```

</details>
