---
title: "compact Quick Start"
output: rmarkdown::html_vignette
description: >
  A minimal end-to-end introduction to compact: from a preprocessed Seurat
  object to a perturbation vector field in under 10 minutes.
vignette: >
  %\VignetteIndexEntry{compact Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

**`compact`** (**co**-expression **m**odule **p**erturbation **a**nalysis for **c**ellular **t**ranscriptomes) is an R package for in-silico gene perturbation analysis in single-cell RNA-seq data. Given a co-expression module, compact silences or activates the module's hub genes, propagates the signal through the gene co-expression network, and computes cell-cell transition probabilities  allowing you to ask: *if this module were knocked down, which cell states would cells move toward?*

This quickstart walks through the minimal workflow: load a dataset, build a neighborhood graph, run one perturbation, and plot the resulting vector field. All key parameters are annotated; everything else is left at its default. For algorithmic depth, multiple modules, and downstream Markov-chain analyses, see the **Basics Simulation** vignette.

> **Prerequisite:** The dataset used here already has co-expression modules computed with `hdWGCNA`. If you need to run hdWGCNA on your own data first, start with the [hdWGCNA tutorial](https://smorabit.github.io/hdWGCNA/articles/basic_tutorial.html).

---

# Section 1: Load Libraries and Data

```{r load-libraries, echo=TRUE}
library(Seurat)
library(tidyverse)
library(cowplot)
library(patchwork)
library(hdWGCNA)
library(compact)

theme_set(theme_cowplot())
set.seed(12345)
```

<details>
<summary>**Dataset generation: simulation and preprocessing steps (click to expand)**</summary>

The dataset was generated from scratch using **splatter** to simulate a linear single-cell trajectory, then processed through Seurat and hdWGCNA. These steps are provided here for full reproducibility; they do not need to be re-run if you are loading the pre-processed object below.

**Step 1: Simulate a linear trajectory with splatter**

```{r simulate-data, eval=FALSE}
library(splatter)
library(scater)

params <- newSplatParams()
params <- setParam(params, "nGenes", 1500)          # 1,500 genes
params <- setParam(params, "batchCells", 500)        # 500 cells
params <- setParam(params, "lib.loc", 8)
params <- setParam(params, "path.nonlinearProb", 0.1)

sim_path <- splatSimulatePaths(
  params,
  group.prob = c(0.2, 0.2, 0.2, 0.2, 0.2),   # 5 equally-sized groups
  de.prob    = 0.75,
  de.facLoc  = 0.2,
  path.from  = c(0, 1, 2, 3, 4),              # linear chain: 1 → 2 → 3 → 4 → 5
  seed       = 12345,
  verbose    = FALSE
)

# remove genes with no zero counts (fully non-sparse genes confound ZINB modeling)
X <- counts(sim_path)
exclude_genes <- names(which(rowMins(X) > 0))
X <- X[setdiff(rownames(X), exclude_genes), ]

seurat_obj <- CreateSeuratObject(
  counts    = X,
  meta.data = as.data.frame(colData(sim_path))
)
```

**Step 2: Standard Seurat preprocessing**

```{r seurat-processing, eval=FALSE}
seurat_obj <- NormalizeData(seurat_obj)
seurat_obj <- FindVariableFeatures(seurat_obj)
seurat_obj <- ScaleData(seurat_obj)
seurat_obj <- RunPCA(seurat_obj)
seurat_obj <- FindNeighbors(seurat_obj, reduction = 'pca', annoy.metric = 'cosine')
seurat_obj <- RunUMAP(seurat_obj, dims = 1:10)
```

**Step 3: Co-expression network analysis with hdWGCNA**

```{r hdwgcna, eval=FALSE}
library(hdWGCNA)
enableWGCNAThreads(nThreads = 8)

# set up hdWGCNA using all genes
seurat_obj <- SetupForWGCNA(
  seurat_obj,
  features   = rownames(seurat_obj),
  wgcna_name = "linear"
)

# construct metacells within each group for robust co-expression estimation
seurat_obj <- MetacellsByGroups(
  seurat_obj      = seurat_obj,
  group.by        = c("Group"),
  reduction       = 'pca',
  k               = 10,
  max_shared      = 5,
  ident.group     = 'Group',
  target_metacells = 100,
  min_cells       = 5
)
seurat_obj <- NormalizeMetacells(seurat_obj)

# fit soft-power threshold and construct the co-expression network
seurat_obj <- SetDatExpr(seurat_obj)
seurat_obj <- TestSoftPowers(seurat_obj)
seurat_obj <- ConstructNetwork(
  seurat_obj,
  soft_power   = 4,
  tom_name     = 'sim_linear',
  overwrite_tom = TRUE
)

# compute module eigengenes and hub gene connectivity scores
seurat_obj <- ModuleEigengenes(seurat_obj)
seurat_obj <- ModuleConnectivity(seurat_obj)

saveRDS(seurat_obj, file = 'data/simulation_linear.rds')
```

</details>

Load the simulated linear-trajectory dataset. This object has been processed through Seurat (normalization, PCA, UMAP) and hdWGCNA (co-expression network, module identification). The co-expression network (TOM) is stored as a separate file; update the path below if your working directory differs.

```{r load-data, eval=TRUE}
# TODO: update this block once this Seurat object is included in the package
seurat_obj <- readRDS('/home/groups/singlecell/smorabito/analysis/COMPACT/data/simulation_linear.rds')

# Update the TOM file path to match your local directory
net <- GetNetworkData(seurat_obj)
net$TOMFiles <- '/home/groups/singlecell/smorabito/analysis/COMPACT/TOM/sim_linear_TOM.rda'
seurat_obj <- SetNetworkData(seurat_obj, net)
```

Visualize the dataset on the PCA embedding, colored by pseudotime group. The five groups (Group1–Group5) represent ordered positions along a single linear differentiation path simulated with **splatter**.

```{r dimplot, eval=TRUE, fig.width=6, fig.height=5}
DimPlot(
  seurat_obj,
  reduction = 'pca',
  group.by  = 'Group',
  label     = TRUE,
  pt.size   = 1.5
) + NoLegend() + coord_equal() 
```

To see which region of the embedding each co-expression module is active in, plot the module eigengenes (hMEs):

```{r module-featureplot, eval=TRUE, fig.width=9, fig.height=3}
plot_list <- ModuleFeaturePlot(
  seurat_obj,
  reduction = 'pca',
  features  = 'MEs',
  order     = TRUE
)
wrap_plots(plot_list, ncol = 3)
```

Note which module shows the strongest gradient from Group1 to Group5  that is the module we will perturb.

---

# Section 2: Build the KNN Graph

compact needs a cell-cell neighborhood graph to compute transition probabilities: transition probabilities are only evaluated between neighboring cells. Run `FindNeighbors` once; all subsequent `ModulePerturbation` calls reuse the same graph.

```{r find-neighbors, eval=TRUE}
seurat_obj <- FindNeighbors(
  seurat_obj,
  reduction    = 'pca',
  dims         = 1:10,
  k.param      = 20,        
  annoy.metric = 'cosine'  
)
```

`FindNeighbors` stores two graphs in the Seurat object: `RNA_nn` (KNN) and `RNA_snn` (shared nearest neighbors). We use the SNN graph (`RNA_snn`) for `ModulePerturbation` in the next section.

> **Tip:** If you ran `FindNeighbors` during standard Seurat preprocessing, you can skip this step and pass the existing graph directly to `ModulePerturbation`. Check which graphs are available with `names(seurat_obj@graphs)`.

---

# Section 3: Run the Perturbation

`ModulePerturbation` is the core function of compact. Internally it:

1. **`ApplyPerturbation`**  modifies expression of the module's top hub genes in the direction specified by `perturb_dir`
2. **`ApplyPropagation`**  diffuses that signal through the gene co-expression network in log-normalized space
3. **`PerturbationTransitions`**  computes cell-cell transition probabilities on the KNN graph

Here we knock down the **blue** module (the module with the highest pseudotime gradient) using `perturb_dir = -1`. The result is stored as a new assay and a new transition probability graph inside the Seurat object.

```{r run-perturbation, eval=TRUE}
seurat_obj <- ModulePerturbation(
  seurat_obj,
  mod              = 'blue',   # co-expression module to perturb
  perturb_dir      = -1,            # negative = knock-down (ZINB mode)
  perturbation_name = 'blue_down',
  graph            = 'RNA_snn',     # neighborhood graph for transition probabilities
  n_hubs           = 10,            # top 10 hub genes receive the primary perturbation
  delta_scale      = 0.2,           # propagation dampening  keep ≤ 0.2 to avoid saturation
  n_iters          = 3              # number of propagation iterations
) 
```

**Expected runtime:** 1–2 minutes on a laptop with a 500-cell dataset.

**Key parameters:**

| Parameter | Value | What it controls |
|---|---|---|
| `perturb_dir` | `-1` | Direction: negative = knock-down, positive = knock-in, `0` = knock-out |
| `n_hubs` | `10` | Number of hub genes to perturb directly; signal diffuses from these to the rest |
| `delta_scale` | `0.2` | Propagation dampening; lower values reduce risk of signal saturation |
| `n_iters` | `3` | Propagation steps through the network |

> **Troubleshooting  graph name mismatch:** If you see an error like *"Graph 'RNA_snn' not found"*, the graph name in your object may differ. Check with `names(seurat_obj@graphs)` and update the `graph=` argument to match.

---

# Section 4: Visualize the Vector Field

`PlotTransitionVectors` projects the cell-cell transition probabilities onto the 2D embedding as a vector field  analogous to an RNA velocity plot. Each arrow summarizes the weighted average direction of transitions within that grid region.

```{r plot-vectors, eval=TRUE, fig.width=6, fig.height=5}
PlotTransitionVectors(
  seurat_obj,
  perturbation_name = 'blue_down',
  reduction         = 'pca',
  color.by          = 'Group',        # color cells by pseudotime group
  plot_mode = 'cells',
  arrow_scale       = 1.5,
  arrow_size = 0.5           
) + ggtitle('blue knock-down') + coord_equal() 
```

**What to look for:** In a linear trajectory, knocking down the module with the highest pseudotime gradient should produce arrows that point *against* the differentiation direction  that is, toward earlier groups (Group1 and Group2) rather than later ones (Group4 and Group5). This confirms that the module drives late-stage differentiation, and removing it would redirect cells toward an earlier state.

> **Tip:** `arrow_scale` controls only the visual arrow length  it does not affect the underlying transition probabilities. Adjust it freely until the arrows are legible. If arrows appear too small, increase it; if they overlap, decrease it.

---

# Section 5: Next Steps

This quickstart covered the minimal compact workflow. The following vignettes build on it:

- **Basics  Simulation** (`simulation_tutorial.Rmd`): full pipeline on a branching trajectory  both perturbation directions, all modules, vector field coherence scoring, and all four Markov-chain analyses (`PredictPerturbationTime`, `PredictAttractors`, `PredictFates`, `PredictCommitment`).

- **Basics  Real Dataset** (`basic_tutorial.Rmd`): compact applied to a real Alzheimer's disease microglia atlas  `AlertSystemScore`, `ComputeDistance`, and `FindShapKeyDriver` for interpretable driver gene prioritization.

- **Advanced Use Cases** (`TF_tutorial.Rmd`): `TFPerturbation` for transcription factor–based perturbations using a regulatory network, and `CustomPerturbation` for arbitrary gene sets.

```{r session-info, eval=TRUE}
sessionInfo()
```
